Typed Functional Genetic Programming

### 2013-08-01 tex/thesis.tex
čtvrtek 6:36

\noindent
%$<$dedication$>$
I would like to express my gratitude to my supervisor RNDr. Petr Pudlák, Ph.D.
for the useful comments, remarks and help with this master thesis.
I would also like to thank my parents, who have supported me throughout
my studies.

@@

Abstrakt:
% abstrakt v rozsahu 80-200 slov; nejedná se však o opis zadání diplomové práce

Klíčová slova:
geneticke programovani, funkcionalni programovani
% 3 až 5 klíčových slov

@@

Abstract:
% abstrakt v rozsahu 80-200 slov v angličtině; nejedná se však o překlad
% zadání diplomové práce

@@

The contribution of this thesis can be seen 
in that it deals with this task using concepts from
theory of typed $\lambda$-calculus and combinatory logic
and shows how to use them in GP. 
Hopefully, it is also shown that those two fields are
nicely linked together.
%, e.g., the concept of \textit{building blocks}
%corresponds to concept of \textit{context}.
Several enhancements are derived from this theoretical background
and shown to have positive effect on performance of our system.  
Our main focus lies in the process of generating the initial population.

@@

Notable highlights of the thesis are: 
\begin{itemize}
 \item \textit{Inhabitation tree} (a slight modification 
   of \textit{inhabitation machine}\cite{barendregt10}
   suited better for reasoning about implementation details) 
   is presented and used for \textit{generation of individuals}.
   This generation process is parameterized by simple 
   \textit{search strategy}. Several search strategies are presented,
   such as \textit{systematic} strategy performing exhaustive enumeration
   of individuals and strategy performing same individual generation 
   as standard GP method \textit{ramped half-and-half}.  
 \item One of those search strategies, the \textit{geometric} strategy, is
   examined in a couple of experiments and shown to have 
   several desirable properties.
 \item The \textit{$\eta$-normalization} of generated individuals
       and \textit{\atTree} representation of individuals, 
       two enhancements of the system inspired from the theoretical
       background of the system are examined and shown to have 
       improving effect on performance of the system.     
 \item \textit{Abstraction elimination} is utilized
   to enable use of simple \textit{mutation operator}.
% \item \textit{Polymorphic} generalization of inhabitation tree,
%   enabling use of polymorphic functions as building blocks, 
%   is presented.

@@

In the field of evolutionary algorithms there is plenty of 
options for selection mechanisms 
for us to choose from. Again we will describe mechanism Koza
used in his first book on GP. It is the \textit{Roulette selection}.
It uses fitness value of each individual in straightforward way to determine probability of selecting this individual.\\
\newpage
Let there be $popSize$ individuals in the population.\\
And let $f_{i}$ be fitness value for individual i 
where $i \in \setDots{1}{popSize}$.

@@

Similarly as for crossover, a maximum permissible depth for mutant individual 
is defined.

@@

Koza in \cite{koza92} uses as default probabilities those values:

@@

\begin{todo}
 & všude nahradit ty množinový za $\Gamma , x:\sigma$ a taky $(x,\Sigma)$ kde se to nehodí.
 & Až to bude dostatečně sepsáno, tak do této kapitoly taky přidat
   část o abstraction elimination a o polymorfním zobecnění inhabitation tree.
   Nezapomenout to pak zmínit i v tomhle úvodu.
\end{todo}

@@

\begin{enumerate}
\item The first reason is rather subjective.
      In my opinion, it simplifies the problem to be
      better graspable. My arguments are following: 
      \begin{enumerate}
      \item
      It reduces the unpleasant fact
      that term generating grammars are not grammars in
      the standard sense. 
      \item
      It brings the problem to 
      its more natural domain, 
      from (textual) grammars 
      to (syntactical) trees.
      \item
      We may think about the whole structure of generating 
      at once since the whole tree, although infinite, is
      fairly simple structure of alternating layers of \textit{and} 
      and \textit{or} nodes. 
      \item
      It enables us to ask such questions
      as: What would be usage of inhabitation trees with 
      number labeled edges coming from \textit{or}-nodes? Et cetera.
      \end{enumerate}   
\item We can think about the inhabitation tree as about 
      some kind of search space. But another view is also 
      possible: We can think about it as about collection
      of typed lambda terms that are generable by that inhabitation
      tree.

      This is even more visible when we introduce
      concept of an \textit{incomplete inhabitation tree}

@@

In the next chapter we will use A*-algorithm together with
inhabitation tree ideas to perform systematic generation and other
kinds of generation of lambda terms;
but maybe even more interesting application of inhabitation trees
involving some above mentioned ideas will be revealed in the 
chapter about future work, since those ideas are not yet implemented.

\section{ Inhabitation machine }

@@

These data are immediately available in the form of
continuously updated graphs and if 
the \textit{fenotype} JavaScript file is present 
for the problem, then is also available immediate 
presentation of the best individual from each generation.\\

@@

@@

\section{Flies and Apples}

@@

\begin{align*}
\sigma = Input &\ar Output\\
\Gamma = \{
dUp              &: Direction                                       ,\\              
dDown            &: Direction                                       ,\\              
dLeft            &: Direction                                       ,\\              
dRight           &: Direction                                       ,\\            
output           &: Move \ar Registers \ar Output                   ,\\            
travel           &: Direction \ar Move                              ,\\          
split            &: Direction \ar Int \ar Registers \ar Move        ,\\       
easySplit        &: Input \ar Move                                  ,\\     
myEnergy         &: Input \ar Int                                   ,\\     
myLastTravel     &: Input \ar Direction                             ,\\                
myWasSuccess     &: Input \ar Bool                                  ,\\      
nAppleDir        &: Input \ar Direction                             ,\\          
nAppleDir        &: Input \ar Direction                             ,\\                    
nAppleDist       &: Input \ar Distance                              ,\\                   
nAppleEnergy     &: Input \ar Int                                   ,\\              
nFlyDir          &: Input \ar Direction                             ,\\                      
nFlyDist         &: Input \ar Distance                              ,\\                    
nFlyEnergy       &: Input \ar Int                                   ,\\              
\end{align*}
\begin{align*}
cAppleDir        &: Input \ar Direction                             ,\\          
cAppleDist       &: Input \ar Distance                              ,\\        
myRegs           &: Input \ar Registers                             ,\\         
xGet             &: Input \ar Int                                   ,\\     
yGet             &: Input \ar Int                                   ,\\     
zGet             &: Input \ar Int                                   ,\\     
dGet             &: Input \ar Direction                             ,\\          
xSet             &: Int \ar Registers \ar Registers                 ,\\          
ySet             &: Int \ar Registers \ar Registers                 ,\\          
zSet             &: Int \ar Registers \ar Registers                 ,\\          
dSet             &: Direction \ar Registers \ar Registers           ,\\                
xInc             &: Registers \ar Registers                         ,\\              
yInc             &: Registers \ar Registers                         ,\\              
zInc             &: Registers \ar Registers                         ,\\              
rotCW            &: Direction \ar Direction                         ,\\                                                             
(==)             &: Int \ar Int \ar Bool                            ,\\                                      
(<=)             &: Int \ar Int \ar Bool                            ,\\               
if'              &: Bool \ar Output \ar Output \ar Output           ,\\                                  
if'              &: Bool \ar Move \ar Move \ar Move                 ,\\                                  
if'              &: Bool \ar Direction \ar Direction \ar Direction  ,\\              
if'              &: Bool \ar Int \ar Int \ar Int                    ,\\     
if'              &: Bool \ar Distance \ar Distance \ar Distance     ,\\        
if'              &: Bool \ar Registers \ar Registers \ar Registers  ,\\                           
0                &: Int                                             ,\\                                     
1                &: Int                                             ,\\                                    
2                &: Int                                             \}\\
\end{align*}

We will not discuss it in great detail, only briefly: First four 
elements are four direction values. Next four functions are
constructors for $Output$ and $Move$ type. Next thirteen functions
are for accessing components of $Input$. Next eleven functions are 
related to $Registers$. The $rotCW$ performs clock-wise rotation of a direction.
Next two functions are comparison functions. Next six are various \textit{if}s. And last 
three are integer constants.

~\\\\\textbf{todo DODĚLAT ASI AŽ POTOM, AŤ SE S TIM MOC NEZDRŽUJU}\\\\

Tedy celkem 44 prvků množiny stavebních bloků, tedy poměrně obrovská množina
na GP poměry.

Takže množina atomických typů je následující:
Move Bool Registers Input Direction Int Output Distance

Používáme tři levely. První level vypadá takle

[obrázek]

Další dva levely jsou reakce na typické chování mouchy 
která používala pro předchozí level triviální strategii
jdi furt za nejbližšim jablkemn. roto jsou následující dva levely 
léčka na tuto strategii.

V druhém levelu se zmíněná moucha nachytá a vleze do 
chodby kde je mnohem mín jablek než v v tom mega jablku z jakblek.

[obrázek]

Třetí dovádí předchozí myšlenku do extrému. 
Moucha se zasekne a nesebere žádné jablko.

V počátečních pokusech byl problém s tím, že mochy s počáteční energii 1
neměli moc tendenci se množit. Proto byla simulace předělána tak, aby
že nyní je počáteční energie 100. To mělo na Mouchy blahodárný vliv.
Ale i pro mouchy sw počátečni energii 1 vznikli některé zajímavé strategie

[povídání o zajímavých strategiích pro poč en 1]

Po navíšení poč energie vznikli tyto zajímavá chování

[povídání o zajímavých strat por poč en 100]\\

@@

\section{Results summary}

@@

The goal of this thesis was to design and implement a system
performing Genetic Programming over some typed functional programming language.
As this typed functional programming language we have chosen the 
\textit{simply typed lambda calculus}.
We have also obligated ourselves to do it in a such way that 
generalizes the standard Genetic Programming, rather than
crates completely new system.

We may say that we have chosen the simplest possible
options in both areas --- the simply typed calculus and the standard GP.
We see this decision as fortunate one --- since it enables us add more
complex features after sufficient exploration of those simplest cases.

@@

A* algorithm has been utilized to drive systematic enumeration
of terms in \lnf.

>>>>

\noindent
%$<$dedication$>$
I would like to express my sincere gratitude to my supervisor RNDr. Petr Pudlák, Ph.D.
for the useful comments, remarks and willingness to help me with 
this thesis.
I would also like to thank my parents, who have supported me throughout
my studies and whole life.

@@

Abstrakt:
% abstrakt v rozsahu 80-200 slov; nejedná se však o opis zadání diplomové práce
{\fontsize{0.35cm}{1em}\selectfont 
V této práci je představen design a implementace
systému řešícího úlohu genetického programování v
simply typed lambda kalkulu. Je zde představena
metoda inicializace počáteční populace založená 
na technice produkující typované lambda termy 
v dlouhé normální formě. Tato metoda je parametrizována
jednoduchou prohledávací strategií. Několik takových
prohledávacích strategií je představeno, jako například
strategie pro systematické generování nebo strategie
odpovídající standardní ramped half-and-half metodě.
Další z představených strategií, strategie jménem 
geometrická strategie je blíže podrobena experimentům,
které ukáží že má několik žádoucích efektů na průběh evoluce, 
jakými jsou zlepšení míry úspěšnosti, nižší časové nároky 
a menší průměrnou velikost termů v porovnání se standardní
ramped half-and-half metodou generování jedinců. Další výkonnostní
zlepšení jsou navržena a podpořena experimenty, jedná se o
$\eta$-normalizaci vygenerovaných jedinců a @-tree
reprezentaci jedinců. Použitý proces eliminace abstrakcí
umožňuje použití jednoduchého podstromy měnícího 
křížení.}

Klíčová slova:
genetické programovaní, funkcionalní programovaní
% 3 až 5 klíčových slov

@@

Abstract:
{\fontsize{0.35cm}{1em}\selectfont 
In this thesis is presented design and implementation
of a system performing genetic programming in simply
typed lambda calculus. Population initialization
method based on term generating technique producing
typed lambda terms in long normal form is introduced. 
This method is parameterized by simple search strategy.
Several search strategies are presented, such as
strategy for systematic generation or strategy
corresponding to standard ramped half-and-half method.
Another such a strategies called \textit{geometric} strategy
is further examined in experiments and shown
to have various desirable effects such as
improved success rate, lesser time consumption and 
smaller average term size in comparison with standard 
ramped half-and-half generating method. Other performance
enhancements are proposed and supported by experiments
such as $\eta$-normalization of generated individuals and 
@-tree representation of individuals. Abstraction elimination
is utilized to enable use of simple tree-swapping crossover.
}

% abstrakt v rozsahu 80-200 slov v angličtině; nejedná se však o překlad
% zadání diplomové práce

@@

%The contribution of this thesis can be seen 
This thesis deals with this task using concepts from
theory of typed $\lambda$-calculus and combinatory logic
and shows how to use them in GP. 
Hopefully, it is also shown that those two fields are
nicely linked together.
%, e.g., the concept of \textit{building blocks}
%corresponds to concept of \textit{context}.
Several enhancements are derived from this theoretical background
and shown to have positive effect on performance of our system.  
Our main focus lies in the process of generating the initial population.

@@

Notable highlights of the thesis are: 
\begin{itemize}
 \item \textit{Inhabitation tree} 
 %(a slight modification 
 %  of \textit{inhabitation machine}\cite{barendregt10}
 %  suited better for reasoning about implementation details) 
   is presented and used for \textit{generation of individuals}.
   This generation process is parameterized by simple 
   \textit{search strategy}. Several search strategies are presented,
   such as \textit{systematic} strategy performing exhaustive enumeration
   of individuals and strategy performing generalization of 
   standard GP term generating method \textit{ramped half-and-half}.  
 \item One of proposed search strategies, the \textit{geometric} strategy, is
   examined in a couple of experiments and shown to have 
   several desirable properties.
 \item The \textit{$\eta$-normalization} of generated individuals
       and \textit{\atTree} representation of individuals, 
       two enhancements of the system inspired from the theoretical
       background are examined and shown to have 
       improving effect on performance of the system.     
 \item \textit{Abstraction elimination} is utilized
   to enable use of simple \textit{mutation operator}.
   Both simple and optimize versions of this transformation are examined.
% \item \textit{Polymorphic} generalization of inhabitation tree,
%   enabling use of polymorphic functions as building blocks, 
%   is presented.

@@

In the field of evolutionary algorithms there is plenty of 
options for selection mechanisms 
for us to choose from. Again we will describe mechanism Koza
used in his first book on GP. It is the \textit{Roulette selection}.
It uses fitness value of each individual in straightforward way to determine probability of selecting this individual.\\
Let there be $popSize$ individuals in the population.\\
And let $f_{i}$ be fitness value for individual i 
where $i \in \setDots{1}{popSize}$.

@@

Analogically as for crossover, a maximum permissible depth for mutant individual 
is defined.

Similarly as Koza in \cite{koza92}, we do not use the mutation operator
in this thesis. This does not mean that I do not like it. 
Conversely, I am looking forward too its enhancing effect in the future work.

@@

Koza\cite{koza92} uses as default probabilities those values:

@@

@@

\begin{enumerate}
\item The first reason is rather subjective.
      In my opinion, it simplifies the problem to be
      better graspable. My arguments are following: 
      \begin{enumerate}
      \item
      It reduces the unpleasant fact
      that term generating grammars are not grammars in
      the standard sense (but instead \textit{2-level-grammars}). 
      \item
      It brings the problem to 
      its more natural domain, 
      from (textual) grammars 
      to (syntactical) trees.
      \item
      We may think about the whole structure of generating 
      at once since the whole tree, although usually infinite, is
      fairly simple structure of alternating layers of \textit{and} 
      and \textit{or} nodes. 
      \item
      It enables us to ask such questions
      as: \textit{What would be usage of inhabitation trees with 
      number labeled edges coming from \textit{or}-nodes?.}
      \end{enumerate}   
\item We can think about the inhabitation tree as about 
      some kind of search space. But another view is also 
      possible: We can think about it as about collection
      of typed lambda terms that are generable by that inhabitation
      tree.

      This is even more visible when we introduce
      concept of an \textit{incomplete inhabitation tree};
      such a tree is created by omission of some subtrees
      with and-node roots (i.e., those accessible by \textit{or} edge).
      \textit{Finite incomplete inhabitation tree} corresponds
      to finite collection of typed \lterms. In very natural 
      sense we can call it a \textit{typed multiterm} --- a collection
      of typed terms forming one tree where those terms share
      the common parts.

      Thus the process of term generation is bidirectional ---
      from (incomplete) inhabitation tree (seen as collection) 
      we may obtain its element by solving it. But we may also 
      take collection of typed trees and compress them
      into one tree by sharing their common parts, which
      happens to be the (incomplete) inhabitation tree 
      that can be solved precisely by those terms which 
      constructed it.

@@

In the next chapter we will use A*-algorithm together with
inhabitation tree ideas to perform systematic generation and other
kinds of generation of \lterms;
but maybe even more interesting application of inhabitation trees
involving some above mentioned ideas will be revealed in the 
chapter about future work, since those ideas are not yet implemented.

\newpage
\section{ Inhabitation machine }

@@

These data are immediately available in the form of
continuously updated graphs and if 
the \textit{phenotype} JavaScript file is present 
for the problem, then is also available immediate 
presentation of the best individual from each generation.
JavaScript functions standing for best individuals are 
also immediately accessible through JavaScript console,
e.g. for debugging purposes.\\

@@

~\\[6em]

\includegraphics[scale=0.4]{skryn.png}

@@

\newpage
\section{Flies and Apples}

@@

\begin{align*}
\sigma = Input &\ar Output\\
\Gamma = \{
dUp              &: Direction                                       ,\\              
dDown            &: Direction                                       ,\\              
dLeft            &: Direction                                       ,\\              
dRight           &: Direction                                       ,\\            
output           &: Move \ar Registers \ar Output                   ,\\            
\end{align*}
\begin{align*}
travel           &: Direction \ar Move                              ,\\          
split            &: Direction \ar Int \ar Registers \ar Move        ,\\       
easySplit        &: Input \ar Move                                  ,\\     
myEnergy         &: Input \ar Int                                   ,\\     
myLastTravel     &: Input \ar Direction                             ,\\                
myWasSuccess     &: Input \ar Bool                                  ,\\      
nAppleDir        &: Input \ar Direction                             ,\\                    
nAppleDist       &: Input \ar Distance                              ,\\                   
nAppleEnergy     &: Input \ar Int                                   ,\\              
nFlyDir          &: Input \ar Direction                             ,\\                      
nFlyDist         &: Input \ar Distance                              ,\\                    
nFlyEnergy       &: Input \ar Int                                   ,\\              
cAppleDir        &: Input \ar Direction                             ,\\          
cAppleDist       &: Input \ar Distance                              ,\\        
myRegs           &: Input \ar Registers                             ,\\         
xGet             &: Input \ar Int                                   ,\\     
yGet             &: Input \ar Int                                   ,\\     
zGet             &: Input \ar Int                                   ,\\     
dGet             &: Input \ar Direction                             ,\\          
xSet             &: Int \ar Registers \ar Registers                 ,\\          
ySet             &: Int \ar Registers \ar Registers                 ,\\          
zSet             &: Int \ar Registers \ar Registers                 ,\\          
dSet             &: Direction \ar Registers \ar Registers           ,\\                
xInc             &: Registers \ar Registers                         ,\\              
yInc             &: Registers \ar Registers                         ,\\              
zInc             &: Registers \ar Registers                         ,\\              
rotCW            &: Direction \ar Direction                         ,\\                                                             
(==)             &: Int \ar Int \ar Bool                            ,\\                                      
(<=)             &: Int \ar Int \ar Bool                            ,\\               
if'              &: Bool \ar Output \ar Output \ar Output           ,\\                                  
if'              &: Bool \ar Move \ar Move \ar Move                 ,\\                                  
if'              &: Bool \ar Direction \ar Direction \ar Direction  ,\\              
if'              &: Bool \ar Int \ar Int \ar Int                    ,\\     
if'              &: Bool \ar Distance \ar Distance \ar Distance     ,\\        
if'              &: Bool \ar Registers \ar Registers \ar Registers  ,\\                           
0                &: Int                                             ,\\                                     
1                &: Int                                             ,\\                                    
2                &: Int                                             \}\\
\end{align*}

\newpage
We will not discuss it in great detail, only briefly: First four 
elements are four direction values. Next four functions are
constructors for $Output$ and $Move$ type. Next twelve functions
are for accessing components of $Input$. Next eleven functions are 
related to $Registers$. The $rotCW$ performs clock-wise rotation of a direction.
Next two functions are comparison functions. Next six are various \textit{if}s. And last 
three are integer constants.

This gives 44 elements of building block context. It is quiet huge 
context, especially if it was a standard GP \TuF.\\

So the set of atomic types is following =\\
$\{Move, Bool, Registers, Input,
Direction, Int, Output, Distance\}$.\\

We currently use three worlds.
First one looks like this:\\

~\\[2em]

\includegraphics[scale=0.8]{fly/1.png}

Other two worlds are reaction on typical behavior 
of a trivial and boring fly which was successful
enough to prevent further evolution.
This boring strategy was just following the path 
of the nearest apple.

Following two worlds are therefore "Devil's gift" for this fly.\\

\includegraphics[scale=0.4]{fly/2.png}
\includegraphics[scale=0.4]{fly/3.png}\\

In the second world the boring fly enters the corridor
and misses the mega apple.
The third world escalates previous idea into absurdity.
The boring fly eats nothing since it dumbly tries to eat the
apple behind the wall.

By introduction of those two new worlds new kinds of behavior emerged,
but not interesting enough to be worth explaining here. One can,
however, check those behaviors out in the gallery of interesting fly 
behaviors, included on the enclosed CD.\\

In the initial experiments there was problem with that flies did not
want to split. This was due to that they were given default energy 1.
Since this problem is not tied by some unchangeable definition,
we changed this default value to 100 and aded a siplified version of split constructor. That had aftermath in new kinds of behavior.

@@

%\section{Results summary}

@@

The goal of this thesis was to design and implement a system
performing GP over some typed functional programming language.
As this typed functional programming language we have chosen the 
\textit{simply typed lambda calculus}.
We have also obligated ourselves to do it in a such way that 
generalizes the standard GP, rather than
crates completely new system.

We may say that we have chosen the simplest possible
option in both areas --- the simply typed calculus and the standard GP.
We see this decision as fortunate one --- since it enables us add more
complex features after sufficient exploration of those simplest cases.

@@

A* algorithm in combination with inhabitation trees
has been utilized to drive systematic enumeration
of typed \lterms in their \lnf. This enumeration was further 
parameterized by simple search strategy in order to enable
such different approaches as \textit{systematic} generation 
and \textit{ramped half-and-half} generation to be definable
in the means of simple search strategy.\\

With this apparatus in hands we introduced novel approach
to term generation by defining the \textit{geometric} search 
strategy. This strategy is further parameterized by parameter
$q$,  but since we wanted to avoid suspicion that success
of this generating method depends on fine-tuning of this
parameter, we used its default value $q = 0.75$ in every experiment.

After being generated all terms undergo process of \textit{abstraction
elimination}, which enables our simple tree swapping crossover operation
which is only genetic operator that we have used in all experiments.

We examined it in three different experiments, which all 
supported the idea that it has very desirable qualities.

First two experiments, \textit{Simple Symbolic Regression} and 
\textit{Artificial Ant}, were performed in order to compare
the \textit{geometric} strategy with the standard GP term 
generating method \textit{ramped half-and-half}.

In both experiments were observed improvements in the following aspects.

\begin{enumerate}
 \item Success rate was improved.
 \item Minimal number of individuals needed to be processed in order to yield 
       correct solution with probability 99\% was lowered.
 \item Run time was significantly reduced. 
 \item Average size of a term was decreased.
\end{enumerate}

In the case of Artificial Ant problem, all improvements were significant.

These results make me believe that \textit{geometric} strategy might
be welcomed reinforcements in the fight against the bogey of 
the GP community --- the \textit{bloat}.
Or at least, it seems to work for those two experiments in a such way.\\

In the third experiment, the \textit{even-parity} problem, geometric strategy 
showed ability to yield correct solution in the initial generation
8 times out of 200 runs (throughout 5 experiments), which is interesting
result since it is uninformed search.

Five consecutive experiments with even-parity problem supported 
hypothesis that $\eta$-normalization of generated terms enhances performance.
We have also seen that use \atTree{}s instead of more traditional \sexprTree{}s
as tree representation of individuals is able to enhanced performance.

Use of optimized abstraction elimination instead of its basic variant 
showed significant improvements in time consumption with little or 
non effects on performance rates.

Implemented system was designed with importance of interactivity in mind,
resulting in server/client architecture for core/GUI components.
